If the Null Fits, You Must Omit: Measurement Artifacts Undermine Group Polarization Replications

### 2025-10-23 GroupPolarizationStatmod/paper/intro-later.tex
tightening intro

\emph{Group polarization} is the name for a form of consensus where
members of a like-minded group become, on average, more extreme in their
opinions after discussing a topic. Group polarization is important
because it may increase social and political tensions if moderates
become more extreme. Decades of studies have replicated detections of
group polarization, typically using an ordinal scale to measure
opinions. However, these studies did not account for ordinal
measurements, which can result in spurious detections of opinion change.
Lacking original data we can still calculate the probability that a
published detection of group polarization was spurious, i.e., to
calculate the \emph{false detection rate}. Across 54 group polarization
experimental conditions from ten representative journal articles, we
calculated false detection rates to be between 0.52 and 0.88, with a
median of 0.75, using a generative model of group polarization
experiments seeded with empirical data. We also use our model to develop
experimental designs that achieve an acceptable false discovery rate.
Much of group polarization research may be unreliable. This analysis can
help change that by enabling others to avoid this hidden pitfall. More
broadly this work demonstrates one important way that replication
success alone does not imply epistemic reliability.

@@

\begin{quote}
In our introductory social psychology course, we have for many years
used the {[}group polarization experimental paradigm{]} as a laboratory
exercise. The exercise works beautifully, but one must be careful to
forewarn a class that {[}group polarization{]} does not occur with every
group\ldots and that the effect is not large (Brown 1986).
\end{quote}

\begin{quote}
One of the most robust findings in social psychology is that of attitude
polarization following discussion with like-minded others (Cooper,
Kelly, and Weaver 2001).
\end{quote}

\begin{quote}
  The good thing about science is that it's true, whether or not you believe
  it (Neil de Grasse Tyson t-shirt, 2010's).
\end{quote}

If an extremist's opinion falls and it is measured with a Likert scale, will
it make a noise? The answer had better be yes if we want to properly measure
opinion change. Accurate measurements of opinion change when tracking
opinions and beliefs over time in surveys and experiments is critical to
design more, sustainable, responsive, and efficient governments,
institutions, and policies (Mason 2018; Klein 2020). One important social
psychology paradigm for explaining opinion change is called \emph{group
polarization}.  Former Obama White House administrator of Information and
Regulatory affairs, and current Harvard Law School professor, Cass Sunstein
writes that the scientific ``law of group polarization''~\cite{Sunstein1999}
provides a ``clue'' to explaining how ``facism\ldots student
radicalism\ldots Islamic terrorism\ldots{[}t{]}he Rwandan
genocide\ldots{[}e{]}thnic Conflict\ldots acts of torture and humiliation by
American soldiers at Abu Ghraib'', and more \cite[p. 1]{Sunstein2009}.

@@

To situate this current paper it will help to introduce some relevant
research traditions and common experimental designs in group polarization,
using the ten studies included in our corpus as touchstones. To begin,
*group polarization* is defined as...(continue basic definition with some
examples from our corpus).

Paragraph dos: Common experimental, measurement, and statistical procedures
to induce and detect group polarization

\subsection{Failure to account for the measurement process can induce
false detections} \label{ssubsec:false-detections}

@@

Group polarization experiments share a common experimental design where,
first, participant opinions are gathered on a topic of interest. Next,
groups are formed that have some initial opinion extremism. Participants
deliberate or discuss the topic for some set time. After the set time,
participants again give their opinion. Group polarization is said to occur
if the mean group opinion became more extreme, i.e., if the magnitude of the
mean group opinion is greater \emph{post-deliberation} compared to
\emph{pre-deliberation}. For example, \cite{Moscovici1969} asked Parisian
high school students to indicate their degree of agreement with the
statement, ``American economic aid is always used for political pressure''.
There were seven options in the selection set, -3 indicating ``strongly
disagree'', +3 indicating ``strongly agree'', and 0 was neutral. Moscovici
and Zavalloni report that the mean pre-deliberation opinion was -0.61 and
the mean post-deliberation opinion was -1.09 (Moscovici and Zavalloni, 1969;
Table 4, p. 132). This is a -0.48 shift to more extreme disagreement with
the statement, reported by Moscovici and Zavalloni as statistically
significant in Table 5 on p. 132. Sometimes group polarization research uses
significance testing for determining group polarization, but Moscovici and
Zavalloni, and others, just report a bare difference in mean opinion. This
alone should make modern researchers think twice before including it and
others as canon, but apparently this has gone unnoticed.

@@

of If we represent the latent distribution

(PREVIOUS TWO PARAGRAPHS SET UP THIS FINAL PARAGRAPH: CONTINUE TO USE
MOSCOVICI 1969 TO EXPLAIN IN THE SIMPLEST TERMS HOW FAILING TO ACCOUNT FOR
MEASUREMENT PROCEDURE CAN LEAD TO FALSE POSITIVES; see Figure
\ref{fig:distros})

>>>>

Echo chambers seem to cause radicalization—and radicalization is risky. Radicalization undermines institutions by reducing their diversity, in turn impairing their \emph{adaptability}.  Social psychologists and others study echo chambers experimentally using \emph{group polarization} experiments. Researchers induce \emph{group polarization} by first asking experiment participants their opinions on some topic, then grouping them together if they agree in spirit, but not degree. The participants then discuss the topic in their groups. Their opinions are then re-measured. Group polarization is when the average group opinion increases. But there’s a fatal problem: participants report opinions on ordinal scales, but this has never been accounted for statistically. This can cause apparent group polarization in observations even if the average latent opinion—the one in people’s heads—doesn’t change, due to ceiling effects. Here we prove that across ten journal articles spanning five decades that 57 of 60 group polarization detections are plausibly false due to this effect—therefore they are not true “replications”. Our method can be used to evaluate any experimental design that measures latent change with ordinal scales. Given how destructive radicalization can be, we must get this right.

@@

If an extremist's opinion falls and it is measured with a Likert scale, will
it make a noise? The answer is often "no"~\cite{Liddell2018,Turner2020}.
This means that a group's opinions can appear to radicalize over time even
when no real shift occurs. The problem matters because Likert and other
ordinal scales are ubiquitous in research on opinion change and
radicalization across the social and cognitive sciences~\cite{Liddell2018}.
One influential theory in this area is \emph{group polarization}—the idea
that discussion among like-minded people pushes opinions toward extremes,
producing ``echo chambers.'' Cass Sunstein, Harvard Law professor
and former White House administrator, has written extensively
about group polarization~\cite{Sunstein2019}. He said group polarization
research provides a ``clue'' to prevent ``radicalism'' and its bad effects
including ``facism\ldots terrorism\ldots genocide\ldots'' and
``torture''~\cite[p. 1]{Sunstein2009}.  Sunstein claims group polarization is
a scientific \emph{law}, a moniker meaning a concrete, quantitative
relationship between variables that will \emph{always be observed as long as
a set of assumptions are met}~\cite{Cartwright1999}. There is reason to be
suspicious first on the fact that Roger Brown admitted in his book
\emph{Social Psychology}, group polarization ``does not occur with every
group'' and ``the effect is not large.'' Given the scope and influence of
group polarization, it is imperative to know whether the hundreds of
``replications'' of group polarization might not really be so because of the
potential they are just \emph{measurement artifacts} due to a mismatch
between the

We are sad to report that

We analyzed sixty representative experimental conditions 
from the group polarization literature to show that the answer is "no"
for nearly all of them  \emph{group polarization}
experiments—\emph{group polarization} is a scientific model of the process by
which \emph{echo chambers} radicalize their members. By \emph{echo chamber} I
mean an online or real-life community where likeminded people discuss 
political or other topics of interest. Group polarization research, then,
could be of great importance both to develop strategies to counteract 
radicalization that undermines our institutions, and to increase support for
sustainable practices. 
Clearly group polarization research has found 
Former Obama White House administrator of Information and
Regulatory affairs, and current Harvard Law School professor, Cass Sunstein

We show the answer is no, if one uses the predominant experimental design
adopted almost Liddell and Kruschke (2018) showed that
this setup can lead to false positives because ordinal scales cannot
track changes in extreme opinions beyond the ceiling of the measurement
scale— In another example
of this experimental design, Burnstein and Vinokur (1975) used a ten-point
scale from 1 to 10 representing “the lowest odds of success acceptable” to
try a certain behavior, where 1 means 10\%—Moscovici and Zavalloni (1969) gave participants a seven-point scale, from “strongly disagree” (-3) to “strongly agree” (+3) to report their opinions. 
reported opinions on the ordinal scale—with discrete numbered bins—were
treated as if they were continuous-valued, or in other words as if they were
the internal mental opinion value itself, whatever that might be
neurobiologically.  In this paper we examine whether group polarization

@@

Following Cartwright’s (1973) critique, social psychologists and others continued work on group polarization, focused on developing and comparing distinct theories of group polarization, with two becoming particularly influential: social comparisons and persuasive arguments. The social comparisons hypothesis says group polarization ultimately occurs due to the human desire to fit in, to hold opinions closer to those around them. Group polarization occurs, then, when a group is composed mostly of extremists, with a sufficient number of people with a weak opinion (Myers and Bishop, 1970; Myers, 1975). Its main “competitor” from the 1970s was the persuasive arguments theory, which basically claimed that rhetoric was the strongest driver of group polarization, not social comparisons (Burnstein and Vinokur, 1973; 1975).

To compare theories, researchers mainly relied on comparing effect sizes or p-values within the context of single idiosyncratic experiments—even after Cartwright implored researchers to develop a coherent theoretical foundation.

For example, Burnstein and Vinokur (1975, p. 420) performed the following experiment and found the results supported their persuasive arguments theory. In one condition participants were only exposed “to the choices of others” but could not discuss—in another they discussed the survey topic with others, but did not know what numerical value they chose. In the final condition, participants could discuss and know each other’s opinions—and this is the only one with a significant p-value indicating group polarization reliably occurred. This alone could not confirm that opinion shifts were due to “persuasive arguments”, so Burnstein and Vinokur also counted how many arguments people made during the discussion sessions. They found that more arguments correlated with greater group polarization shifts to more extreme opinions, —confirmation that persuasive arguments beats social comparisons, according to them. About a decade later, Isenberg (1986) performed an aggregated “significance test” based on reported p-values, comparing the aggregate across papers claiming to prove the persuasive arguments hypothesis versus the social comparisons hypothesis.

Additional hypotheses about what causes group polarization proliferated at the end of the last century, despite zero progress towards commensurability or even awareness of the problem. Other notable hypotheses include the self-categorization and social decisions scheme hypotheses. Self-categorization augmented the social comparisons hypothesis with a consideration of idiosyncratic beliefs people hold that are the result of their own personal lived experience (Abrams, et al., 1990; Hogg, et al., 1990). The social decisions hypothesis posits that contextual factors like social network structure could be at least as important as one’s opinions and the opinions of others . Noah Friedkin, a sociologist, notably published perhaps the most rigorous study on group polarization, adopting the social decisions scheme hypothesis, predicting that social connectivity within a group could amplify group polarization (Friedkin, 1999).

\subsection{\mt{CUT??} Failure to account for the measurement process can induce
false detections} \label{ssubsec:false-detections}

@@

Group polarization experiments share a common experimental design where,
first, participant opinions are gathered on a topic of interest. Next,
groups are formed that have some initial opinion extremism. Participants
deliberate or discuss the topic for some set time. After the set time,
participants again give their opinion. Group polarization is said to occur
if the mean group opinion became more extreme, i.e., if the magnitude of the
mean group opinion is greater \emph{post-deliberation} compared to
\emph{pre-deliberation}. For example, \cite{Moscovici1969} asked Parisian
high school students to indicate their degree of agreement with the
statement, ``American economic aid is always used for political pressure''.
There were seven options in the selection set, -3 indicating ``strongly
disagree'', +3 indicating ``strongly agree'', and 0 was neutral. Moscovici
and Zavalloni report that the mean pre-deliberation opinion was -0.61 and
the mean post-deliberation opinion was -1.09 (Moscovici and Zavalloni, 1969;
Table 4, p. 132). This is a -0.48 shift to more extreme disagreement with
the statement, reported by Moscovici and Zavalloni as statistically
significant in Table 5 on p. 132.

@@

### 2025-10-23 GroupPolarizationStatmod/paper/intro-latest.tex
tightening intro

\emph{Group polarization} is the name for a form of consensus where
members of a like-minded group become, on average, more extreme in their
opinions after discussing a topic. Group polarization is important
because it may increase social and political tensions if moderates
become more extreme. Decades of studies have replicated detections of
group polarization, typically using an ordinal scale to measure
opinions. However, these studies did not account for ordinal
measurements, which can result in spurious detections of opinion change.
Lacking original data we can still calculate the probability that a
published detection of group polarization was spurious, i.e., to
calculate the \emph{false detection rate}. Across 54 group polarization
experimental conditions from ten representative journal articles, we
calculated false detection rates to be between 0.52 and 0.88, with a
median of 0.75, using a generative model of group polarization
experiments seeded with empirical data. We also use our model to develop
experimental designs that achieve an acceptable false discovery rate.
Much of group polarization research may be unreliable. This analysis can
help change that by enabling others to avoid this hidden pitfall. More
broadly this work demonstrates one important way that replication
success alone does not imply epistemic reliability.

@@

\begin{quote}
In our introductory social psychology course, we have for many years
used the {[}group polarization experimental paradigm{]} as a laboratory
exercise. The exercise works beautifully, but one must be careful to
forewarn a class that {[}group polarization{]} does not occur with every
group\ldots and that the effect is not large (Brown 1986).
\end{quote}

\begin{quote}
One of the most robust findings in social psychology is that of attitude
polarization following discussion with like-minded others (Cooper,
Kelly, and Weaver 2001).
\end{quote}

If an extremist's opinion falls and it is measured with a Likert scale,
will it make a noise? The answer had better be yes if we want to
properly measure opinion change. Accurate measurements are critical for
developing rigorous and reliable methods to promote more stable,
sustainable, responsive, and efficient governments and institutions
(Mason 2018; Klein 2020). Unfortunately, social and behavioral
scientists have often used statistical methods that distort opinion
change when it's measured on an ordinal, Likert-style scale (e.g.~-3
indicates ``Strongly disagree'', +3 indicates ``Strongly agree'', and 0
is neutral) (Liddell and Kruschke 2018). This includes a phenomenon
called \emph{group polarization} that former Obama White House official
and law professor Cass Sunstein relied on to explain why ``people become
extremists'' and why ``political and cultural polarization'' is ``so
pervasive in America'', as the publisher summarizes (Sunstein 2009).
Sunstein suggests group polarization research provides a ``clue'' to
explaining how ``facism\ldots student radicalism\ldots Islamic
terrorism\ldots{[}t{]}he Rwandan genocide\ldots{[}e{]}thnic
Conflict\ldots acts of torture and humiliation by American soldiers at
Abu Ghraib'', and more (p.~1).

% \begin{itemize}
% \tightlist
% \item
We developed a generative model of opinion change to identify, if possible,
null opinion change models that could spuriously appear like group polarization
when naively measured using a Likert scale without accounting for the measurement procedure. 
We further applied our method to estimate the family-wise error rate of
detections of group polarization in a corpus of ten influential papers from 
across several decades and from various theoretical traditions. We turn this criticism of past
blunders constructive by providing a procedure to calculate what Cohen's $d$ should be
used to guarantee a family-wise error rate (i.e., Type I error rate) of 5\% or less.

We calculate the severity of the problem in two steps. In the first step we 
provide proof by example that there exists a null opinion change model in terms of \emph{latent, continuous}
psychological opinions that looks like group polarization when measured via participant behavior, i.e.,
by their response on an \emph{ordinal survey scale}. If such a null model is found, this means that we cannot
rule out a null effect, given the published data. This means, therefore, that the published detection is
just as likely false as it is true, so we say it is a \emph{plausibly false detection of group polarization},
or a \emph{plausibly spurious detection} for short.

In the second step of the analysis of whether, we simulate observations
of group polarization experiments where participant opinions are drawn from the 
null model distributions identified in the first step,
then these are measured on a Likert scale for simulated group polarization survey
responses on an ordinal scale. We then calculate the
family-wise error rate (i.e., Type I error rate) and false detection rate using
various effect sizes to determine significance of the effect in terms of Cohen's $d$.
The family-wise error rate is defined as the probability that a known null 
effect is detected as significant. We use this to then estimate the *false detection rate*, which is the
probability that a detection of an effect is a false one, which depends on the relative prevalence of
true and false positive detections in the literature.
Finally, we put our model to constructive use to show what Cohen's $d$ must be set to
to limit the family-wise error rate to 5\%, denoted $d*$. This will not help rescue any of
the published results here found to be plausibly spurious. But, it provides a measure of the weakness of
the published experimental designs independent of their shared fatal failure to 
account for the measurement procedure: greater $d*$ indicates weaker designs.

We find that 95\% (54 of 57) of group polarization detections are plausibly spurious.
In other words, empirical support for group polarization has been seriously undermined
just by removing these results from the count of group polarization replications.
There is no reason to expect other papers to be more reliable. (Summarize other
two findings, 1. FWER/FDR and 2. $d*$ for FWER < 5\%)

Our methods can be turned on any data-driven social science analysis of
data where participants report a latent, subjective psychological
variable (mood, physiological state, etc.), and where various theories
supposedly battle it out to best represent the phenomenon of interest
via null-hypothesis significance testing using ordinal data as continuous. 
This may just be the first
of many bricks to fall since \emph{all 68} of the articles that
mentioned the word ``Likert'' and analyzed ordinal data used metric
models to detect latent psychological variable change across three
influential psychology journals, \emph{Psychological Science}, the
\emph{Journal of Experimental Psychology: General}, and the
\emph{Journal of Personality and Social Psychology} (Liddell and
Kruschke 2018). This further confirms that replication is far from the only
foundational crisis that behavioral science faces\cite{Yarkoni2022,Turner2022}.

Replication is not enough:
strong, mechanistic models must be developed to be able to predict Type I and II
error rates \emph{ahead of time}, as we demonstrate here. We only demonstrate how to
calculate expected Type I error rates for a measurement design, i.e., the rate of false 
positives. Low statistical power exacerbates false detection rates. So, to be confident in replication,
a pre-registered measurement design should use this style of simulation to 
estimate the statistical power. Tangled, weak, informal
verbal models of complex psychological and social phenomena like those that motivated
empirical studies of group polarization tend to be vulnerable
to stastical weaknesses. It is difficult to incorporate prior knowledge such
as experiment structure without a sufficiently rigorous and appropriate 
formal system \cite{Smaldino2017,Smaldino2020,Flake2020,Deffner2024}.

@@

(DO THE MINIMUM AFTER EVERYTHING ELSE)

To situate this current paper it will help to introduce some relevant research
traditions and common experimental designs in group polarization, using the ten
studies included in our corpus as touchstones. To begin, *group polarization*
is defined as...(continue basic definition with some examples from our corpus).

Paragraph dos: 
Common experimental, measurement, and statistical procedures to induce and
detect group polarization

\subsection{How study design can induce false detections of group polarization}
\label{ssubsec:false-detections}

An opinion is never measured directly, only a behavior that indicates an opinion
in a behavioral experiment or survey.  For example, the location of a mark
that a participant makes is measured, not the complex neural activity that
underlies forming opinions.
Theoretically, an internal opinion in the form of neural activity is a
\textit{latent opinion}, from the Latin \emph{latentem} meaning ``lie
hidden''.  Participants report \text{observed opinions}, typically as one of
a finite number of ranked, successive integer values that form the
\emph{selection set}. The minimum possible value in the selection set might
mean ``strongly disagree'' and the maximum value represents ``strongly
agree''.

Group polarization experiments share a common experimental design where,
first, participant opinions are gathered on a topic of interest. Next, groups
are formed that have some initial opinion extremism. Participants deliberate
or discuss the topic for some set time. After the set time, participants
again give their opinion. Group polarization is said to occur if the
mean group opinion became more extreme, i.e., if the magnitude of the mean
group opinion is greater \emph{post-deliberation} compared to
\emph{pre-deliberation}. For example,
\cite{Moscovici1969} asked Parisian high school students to indicate their
degree of agreement with the statement, ``American economic aid is always
used for political pressure''. There
were seven options in the selection set, -3 indicating ``strongly disagree'',
+3 indicating ``strongly agree'', and 0 was neutral. Moscovici and Zavalloni
report that the mean pre-deliberation opinion was -0.61 and the mean
post-deliberation opinion was -1.09 (Moscovici and Zavalloni, 1969; Table 4,
p. 132). This is a -0.48 shift to more extreme disagreement with the 
statement, reported by Moscovici and Zavalloni as statistically significant
in Table 5 on p. 132.

(PREVIOUS TWO PARAGRAPHS SET UP THIS FINAL PARAGRAPH: CONTINUE TO USE MOSCOVICI 1969 TO
EXPLAIN IN THE SIMPLEST TERMS HOW FAILING TO ACCOUNT FOR MEASUREMENT PROCEDURE
CAN LEAD TO FALSE POSITIVES; see Figure \ref{fig:distros})

\subsection{Overview}

In the next section, we introduce the various formal and computational Methods 
used to obtain our findings outlined above. We then explain our findings in detail in the
Analysis section. We close with a Discussion of how this is not a defeat, but
a new beginning for group polariation, given that there are good reasons to expect
group polarization to occur. We also explain how to expanding this analysis to
examine other published studies on group polarization and across social and behavioral
science disciplines.

>>>>

Echo chambers seem to cause radicalization—and radicalization is risky.
Radicalization undermines institutions by reducing their diversity, 
which in turn impairs their adaptability.  Social psychologists and others study
echo chambers experimentally using \emph{group polarization} experiments.
Researchers induce \emph{group polarization} by first asking experiment
participants their opinions on some topic, then grouping them together if
they agree in spirit, but not degree. The participants then discuss the topic
in their groups. Their opinions are then re-measured. Group polarization is
when the average group opinion increases. But there’s a fatal problem:
participants report opinions on ordinal scales, but this has never been
accounted for statistically. This can cause apparent group polarization in
observations even if the average latent opinion—the one in people’s
heads—doesn’t change, due to ceiling effects. Here we prove that across ten
journal articles spanning five decades that 54 of 57 reported polarization effects are plausibly false due to this artifact, and thus cannot be counted as replications. Our method can be used to evaluate any experimental design that measures latent change with ordinal scales. Given how destructive radicalization can be, we must get this right.

@@

If an extremist’s opinion falls and it is measured with a Likert scale, will
it make a noise? The answer is often no~\cite{Liddell2018,Turner2020}. This
means a group’s opinions can appear to radicalize over time even when no real
shift occurs. The problem matters because Likert and other ordinal scales are
ubiquitous in research on opinion change and radicalization across the social
and cognitive sciences~\cite{Liddell2018}. One influential framework in this
area is group polarization—the idea that discussion among like-minded people
pushes opinions toward extremes, producing ``echo chambers.'' Cass Sunstein, a
Harvard Law professor and former White House administrator, has written
extensively on the topic~\cite{Sunstein2019}, claiming that group
polarization is a scientific law—a regular, quantitative relationship that
will always be observed when certain assumptions are
met~\cite{Cartwright1999}. Yet Roger Brown himself noted in Social Psychology
that group polarization ``does not occur with every group'' and that ``the
effect is not large.'' Given the influence of this theory and the hundreds of
claimed ``replications,'' it is essential to ask whether many of these findings
may instead be measurement artifacts—spurious effects created when the
``neutral'' point of the measurement scale fails to align with the group’s own
center of opinion.

The group polarization hypothesis is that discussion within a group of
likeminded people causes the group's average opinion to increase, or
\emph{polarize}~\cite{Brown1986,Brown2000}.
Group polarization experiments therefore attempt to
polarize groups as follows: first, participants report their opinion on
some topic or prompt chosen by the experimenters in advance. Second, 
participants are put into likeminded groups to
discuss the topic. Following discussion, participants again report their
opinions on the topic. Group polarization is defined as an increase in
the average opinion following discussion. If the group polarized, and we 
assume that people in the group found consensus—meaning their opinions agreed
more after the discussion than before—then initially moderate opinions must have
increased more than more extreme ones.

We are sad to report that 95\% of experimental reports of group polarization
across sixty published experimental conditions cannot be counted as
replications. We explain here how evidence of group polarization is undermined
by a basic measurement flaw that obscures opinion change among extremists.
This opens the possibility that what seems to be group polarization could just
be \emph{mere agreement}, with no change in mean opinion, if only increases in
opinion were measured but not decreases.  To evaluate whether the empirical
record reflects genuine opinion change or merely measurement artifacts, I
simulated published experiments under conditions of mere agreement. Each
simulation reproduced the core elements of published designs—pre-discussion
survey, group interaction, post-discussion survey—while holding the latent
mean opinion constant. Because the original datasets are lost and published
results rarely provide enough information to reconstruct them, we cannot know
whether any reported effect reflects a real shift or a measurement artifact.
When distinct latent processes can generate indistinguishable results, the
effect is non-identifiable—and no claim of replication can stand.

A relatively simple step could fix the problem: re-center the post-discussion
scale on each group’s own pre-discussion mean, aligning measurement with the
group’s frame rather than an arbitrary external ``neutral.'' This adjustment
restores the basic condition for inference: measuring change within a
consistent frame of reference.

But even if this measurement problem is fixed,
it is one of several severe problems in this kind of social science: Unstable
opinions, unmodeled input variance, and incommensurate experimental designs
all uniquely undermine any cumulative insight group polarization might
claim—problems known about and ignored for decades~\cite{Cartwright1973}. No
degree of scientific lawfulness can be established without evidence gathered
from rigorously modeled, commensurate experiments~\cite{Cartwright1999}.

@@

Group polarization research began in the 1960s, which grew rapidly with
funding from the National Institutes of Mental Health (NIMH), and widespread interest
in what was then a surprising new science of radicalization.
Dorian Cartwright (1973) reviewed the first NIMH grants and at the time said,
`` ''.
